Probabilistic Forecasting of Geomagnetic Indices using Gaussian Process Models

### 2017-08-16 revised-chapter.tex
Revisions

\begin{document}

@@

%Talk about NARMAX Dst
Important empirical geomagnetic prediction models include the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology (see \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN) based models (\citet{Lund}, \citet{JGRA:JGRA17461}, \citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series prediction of $Dst$ and $Kp$ indices from interplanetary magnetic field data and solar wind parameters.

\subsection{Probabilistic Forecasting}
%Talk about need for probabilistic forecasts.ny
Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$. One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model, using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.

@@

\item \textbf{Model Order Selection}: In order to find appropriate values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$, solar wind speed and IMF $B_z$ respectively, create models upto a certain maximum time lag i.e. $3 \leq p \leq 8$ for GP-AR and $3 \leq p + p_v + p_b \leq 8$ for GP-ARX. For each such model perform the model selection routine to select values of the kernel hyperparameters and evaluate the resultant model on the validation set. Choose the value of $p, p_v, p_b$ which maximises predictive performance on the validation set.

@@

After running the workflow steps as outlined in section \ref{sec:workflow} above, we can analyse the performance of the GP-AR and GP-ARX family of models on task of OSA $Dst$ prediction in the following series of charts.

@@

@@

From the validation results, we can choose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$. Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.

>>>>

\begin{document}
\maketitle

@@

% Talk about NARMAX Dst
Important empirical geomagnetic prediction models include the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology (see \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN) based models (\citet{Lund}, \citet{JGRA:JGRA17461}, \citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series prediction of $Dst$ and $Kp$ indices from interplanetary magnetic field data and solar wind parameters.

\subsection{Probabilistic Forecasting}
% Talk about need for probabilistic forecasts.ny
Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$. One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model, using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.

@@

\item \textbf{Model Order Selection}: In order to find appropriate
  values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$,
  solar wind speed and IMF $B_z$ respectively, create models upto a
  certain maximum time lag i.e. $3 \leq p \leq 12$ for GP-AR and $3
  \leq p + p_v + p_b \leq 12$ for GP-ARX. For each such model perform
  the model selection routine to select values of the kernel
  hyperparameters and evaluate the resultant model on the validation
  set. Choose the value of $p, p_v, p_b$ which achieves the predictive
  performance on the validation set (i.e. minimizes the \emph{mean
    absolute error}).

@@

After running the workflow steps as outlined in section
\ref{sec:workflow} above, we are at a position to analyse the
performance of the GP-AR and GP-ARX family of models on task of OSA
$Dst$ prediction.

@@

The model validation workflow described in section \ref{sec:workflow}
produces data containing model validation scores for GP-AR and GP-ARX
models with different values of autoregressive orders $p_t$. In order
to better understand the over all trend, we group the performance
scrores by unique values of $p_t = p + p_v + p_b$ and analyse the
summary statistics with increasing $p_t$

@@

From the charts above, we can choose the model order which yields the
best \emph{mean absolute error}, for GP-AR it is $p_t = 6$ while for
GP-ARX it is $p_t = 11$. Further examination of the validation results
shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
